---
title: "Merging and Processing CoreMS Outputs"
author: "RED"
date: "2025-02-12"
output: html_document
---

The purpose of this markdown is to merge outputs generated using our CoreMS python script, and run that merged output through ftmsRanalysis.

To run this full markdown, you'll need to:

1) Install R + Rstudio (https://posit.co/download/rstudio-desktop/)
2) Install necessary packages - install.packages(c(“devtools”, “tidyverse”, “easycsv”))
3) Install ftmsRanalysis - devtools::install_github("EMSL-Computing/ftmsRanalysis")

## Load packages
This section loads all of the required R packages for this pipeline and allows you to tweak the import settings. *Be sure to specify your dataset name.* The script should prompt you for your input directory.

```{r setup, include=T, warning=F}

# ############### #
### User Inputs ###
# ############### #

# name of the dataset
dataset.name = "Test_Processed"

# merge by formula or calibrated m/z
merge.by = "formula"

# select formula selection (e.g., best match or all matches)
formula.select = "best"

# full path to directory with CoreMS output
path_to_dir = easycsv::choose_dir()

# select output name
output_dir = paste0(path_to_dir, "/Merged_Output/")

# ################# #
### Load Packages ###
# ################# #

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = path_to_dir)
require(tidyverse)
require(ftmsRanalysis) # https://github.com/EMSL-Computing/ftmsRanalysis

# catching internet activity
is_online <- function(site="https://www.google.com") {
  tryCatch({
    readLines(site,n=1)
    TRUE
  },
  warning = function(w) invokeRestart("muffleWarning"),
  error = function(e) FALSE)
} # pulled from Stack Overflow

# lambda functions 
# this is configured to pull the function from our repo, but has a fail safe that
# you can configure if you lose an internet connection
# Modified from https://github.com/hyunseobsong/lambda with permission
if(is_online()){
  source("https://raw.githubusercontent.com/danczakre/RCSFA_CoreMS/refs/heads/main/getLambda.R")
} else {
  source("path/to/getLambda.R")
}

###### DO NOT EDIT BELOW THIS LINE IN THIS CHUNK ######

```

## Importing data into R
This chunk loads each corems.csv file generated using our CoreMS script. This will navigate to the folder that you indicated above and load in every single file; there is no mechanism to control for diverse datasets.

```{r import}

# ###################################### #
### Import processed .csvs from CoreMS ###
# ###################################### #

# Change directory
setwd(path_to_dir)

# list CoreMS output files
file.list = list.files(pattern = "corems.csv")

# empty objects
mol = NULL
data = NULL

# loop through files and import
for(file in file.list){
  
  ### Loading and cleaning
  # load CoreMS ouput
  temp = read.csv(file = file, check.names = F)
  
  # fix molecular formulas
  temp$`Molecular Formula` = gsub(" ", "", temp$`Molecular Formula`)
  
  # check elemental columns, adding if necessary
  if(!"N" %in% names(temp)) temp = temp %>% add_column(N = NA)
  if(!"S" %in% names(temp)) temp = temp %>% add_column(S = NA)
  if(!"P" %in% names(temp)) temp = temp %>% add_column(P = NA)
  
  # remove unassigned peaks if merging by formulas
  if(merge.by == "formula"){
    if(length(which(temp$`Heteroatom Class` %in% "unassigned")) == 0){
      print(paste0("No unassigned formulas present in ", file))
    } else {
      temp = temp[-which(temp$`Heteroatom Class` %in% "unassigned"),]
    }
  }
  
  # storing sample name
  samp.name = gsub(".csv", "", file)
  
  
  ### Filtering by predefined rules
  # dropping formulas with an O:P ratio > 4 (the filter is not working in CoreMS)
  op.ratio = temp$O/temp$P
  
  if(length(which(op.ratio < 4)) > 0){
    temp = temp[-which(op.ratio < 4),]
  }
  
  # filtering by NSP <=4
  nsp.sum = apply(temp[,c("N", "S", "P")], 1, function(x) sum(x, na.rm = T))
  
  if(length(which(nsp.sum > 4)) > 0){
    temp = temp[-which(nsp.sum > 4),]
  }
  
  # selecting formulas
  if(formula.select == "best" | merge.by == "mass"){
    # Selecting highest confidence score
    temp = temp %>% group_by(`Calibrated m/z`)%>%
      slice_max(`Confidence Score`) %>% ungroup()
  }
  
  # pre-filtering duplicates
  if(merge.by == "formula"){
    temp = temp %>% group_by(`Molecular Formula`)%>%
      slice_max(`Peak Height`) %>% ungroup()
  }
  
  ### Organization and combination
  # separating molecular data
  temp.mol = temp %>%
    select(c("Calibrated m/z", "C", "H", "O", "N", "S", "P", "Molecular Formula",
             "Is Isotopologue", "Heteroatom Class", "Calculated m/z", 
             "m/z Error (ppm)"))
  
  # separating sample data
  if(merge.by == "mass"){
    temp.data = temp %>%
      select(c("Calibrated m/z", "Peak Height")) %>%
      rename(!!sym(samp.name) := `Peak Height`) # this !!sym notation is necessary to convert variable to dplyr object
  } else if(merge.by == "formula"){
    temp.data = temp %>%
      select(c("Molecular Formula", "Peak Height")) %>%
      rename(!!sym(samp.name) := `Peak Height`)
  }
  
  # merging data
  if(is.null(data)){
    data = temp.data
    mol = temp.mol
  } else {
    if(merge.by == "mass"){
      data = data %>% full_join(temp.data, by = "Calibrated m/z") %>%
        distinct() # full joins so we don't lose data
      mol = rbind(mol, temp.mol)
      mol = mol[!duplicated(mol$`Calibrated m/z`),]
    } else if(merge.by == "formula"){
      data = data %>% full_join(temp.data, by = "Molecular Formula") %>%
        distinct()
      mol = rbind(mol, temp.mol)
      mol = mol[!duplicated(mol$`Molecular Formula`),]
    }
  }
  
  print(paste(samp.name, (nrow(temp.data)-nrow(temp.mol)), (nrow(data)-nrow(mol))))
  print(paste0("Finished processing ", samp.name, " which is #", which(file.list %in% file), " out of ", length(file.list), " files (", round((which(file.list %in% file)/length(file.list))*100, digits = 2), "%)"))
  
}

# resolving duplicated formula assignments
if(merge.by == "formula"){
  # Occasionally, two masses are assigned the identical formula but have 
  # different "abundances this will simply aggregate them by average
  
  # The Mol object should automatically have this resolved via the !duplicated
  # step
  
  data = data %>%
    group_by(`Molecular Formula`) %>%
    summarise_all(.funs = "mean")
}

# ensuring the two data files match
if(merge.by == "mass"){
  data = data[order(data$`Calibrated m/z`),] # easiest way to ensure correct order to to let R handle ordering
  mol = mol[order(mol$`Calibrated m/z`),]
} else if(merge.by == "formula"){
  data = data[order(data$`Molecular Formula`),]
  mol = mol[order(mol$`Molecular Formula`),]
  
  if(!identical(data$`Molecular Formula`, mol$`Molecular Formula`)){
    stop("Something unpredictable went very wrong - your molecular formula do not match across the two files necessary for ftmsRanalysis. Good news - all the .csvs *should* still be saved.")
  }
}

# creating output directory if it doesn't exist
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}

# writing merged dataset
write.csv(data, paste0(output_dir, dataset.name, "-Unprocssed_Data.csv"), 
          quote = F, row.names = F)
write.csv(mol, paste0(output_dir, dataset.name, "-Unprocessed_Mol.csv"), 
          quote = F, row.names = F)

```

## Running ftmsRanalysis
This chunk of code runs reconfigures the data to better mesh with ftmsRanalysis and then runs through a few of the capabilities built into ftmsRanalysis. Specifically, we will be 1) removing anything with an isotopic signature, 2) filtering any masses below 150 m/z or above 900 m/z, 3) calculating various derived metrics like AI and NOSC, and 4) assigning molecular formula to various boundary sets.

```{r ftmsRanalysis}

# ##################################### #
### Preparing files for ftmsRanalysis ###
# ##################################### #

# if merged by formula, re-adding mass information
if(merge.by == "formula"){
  data = data.frame(`Calibrated m/z` = mol$`Calibrated m/z`, 
                    data[,-which(colnames(data) %in% "Molecular Formula")],
                    check.names = F)
} 

# filling NA's with 0's
data[is.na(data)] = 0
mol$N[is.na(mol$N)] = 0
mol$S[is.na(mol$S)] = 0
mol$P[is.na(mol$P)] = 0

# ensuring elements are numeric
mol$C = as.numeric(mol$C)
mol$H = as.numeric(mol$H)
mol$O = as.numeric(mol$O)
mol$N = as.numeric(mol$N)
mol$S = as.numeric(mol$S)
mol$P = as.numeric(mol$P)

# ensuring objects are dataframes (tibbles are the enemy here)
data = as.data.frame(data)
mol = as.data.frame(mol)

# creating empty factor object
factor = data.frame(Sample_ID = colnames(data)[-1], Location = "Somewhere")

# if some duplicated masses got through - picking the assignment with lowest error
dup.mass = mol$`Calibrated m/z`[duplicated(mol$`Calibrated m/z`)] # identify duplicated masses

if(!length(dup.mass) == 0){
  dup.ind = which(mol$`Calibrated m/z` %in% dup.mass) # find indices for duplicated masses
  dup.filt = data.frame(index = dup.ind, mol[dup.ind,], check.names = F) # create mol object with indices and mol
  
  dup.filt = dup.filt %>% group_by(`Calibrated m/z`) %>% 
    slice_max(abs(`m/z Error (ppm)`)) # identify highest error
  
  data = data[-dup.filt$index,] # remove the highest error dups from the dataset
  mol = mol[-dup.filt$index,]
}


if(!identical(data$`Calibrated m/z`, mol$`Calibrated m/z`)){
  stop("Something has gone tragically wrong - time to troubleshoot with little to no information!")
} # quick identity check

# ######################### #
### Running ftmsRanalysis ###
# ######################### #

# convert to ftmsRanalysis object
peak_icr = as.peakData(e_data = as.data.frame(data), f_data = factor, e_meta = as.data.frame(mol), 
                       edata_cname = "Calibrated m/z", mass_cname = "Calibrated m/z", fdata_cname = "Sample_ID", 
                       c_cname = "C", h_cname = "H", o_cname = "O", n_cname = "N", s_cname = "S", p_cname = "P", 
                       isotopic_cname = "Is Isotopologue", isotopic_notation = "1")

# calculating derived metrics (e.g., NOSC, DBE, etc.)
peak_icr = compound_calcs(peak_icr)

# assigning boundary sets
peak_icr = assign_class(peak_icr, boundary_set = "bs1")
peak_icr = assign_class(peak_icr, boundary_set = "bs2")
peak_icr = assign_class(peak_icr, boundary_set = "bs3")

# filter by mass
filter_obj = mass_filter(peak_icr)
peak_icr = applyFilt(filter_obj, peak_icr, min_mass = 100, max_mass = 1000)

# lambda magic
get_comp = get_compositions(peak_icr$e_meta)
lambda = as.data.frame(get_lambda(get_comp$chemical_compositions))
names <- rep("", 62)

names[1:12] <- c("delGcox0","delGd0","delGcat0","delGan0","delGdis0","lambda0",
                 "delGcox","delGd","delGcat","delGan","delGdis","lambda")

stoich_colnames <- c("donor","h2o","hco3","nh4","hpo4","hs","h","e","acceptor",
                     "biom")
stoich_types <- c("stoichD","stoichA","stoichCat","stoichAn","stoichMet")

for (i in 1:length(stoich_types)) {
  names[((i-1)*10+13):(i*10+12)] <- array(sapply(stoich_types[i], paste, 
                                                 stoich_colnames, sep="_"))
}
colnames(lambda) <- names
lambda['MolForm'] <- get_comp$formulas
lambda = as.data.frame(lambda[,c("MolForm", "delGcox0", "delGcox", "lambda0", 
                                 "lambda", "delGd0", "delGd")])
colnames(lambda) = c("MolForm","delGcox0PerCmol","delGcoxPerCmol", "lamO20",
                     "lamO2","delGd0","delGd")
lambda = lambda[!duplicated(lambda$MolForm),]

# adding lambda information into mol object
peak_icr$e_meta = peak_icr$e_meta %>% left_join(lambda, by = "MolForm")

# Write results
write.csv(peak_icr$e_data, 
          paste0(output_dir, dataset.name, "-Processed_Data.csv"), 
          quote = F, row.names = F)
write.csv(peak_icr$e_meta, 
          paste0(output_dir, dataset.name, "-Processed_Mol.csv"), 
          quote = F, row.names = F)

```