---
title: "smet_results_analysis"
author: "Matt Kaufman and Vanessa Garayburu-Caruso"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) # Sets working Directory to the R project

```

##Matt Kaufman and Vanessa Garayburu-Caruso, vanessa.garayburu-caruso@pnnl.gov, Pacific Northwest National Laboratory

This code assembles average values for ecosystem respiration, gross primary production, and K600 across the deployment period for each of the SSS sites into one summary file. It also calculates a period-average depth and uses that to compute per-volume rates from the per-area rates that stream metabolizer produces.

##import GPP and ER daily means from SM results and combine
```{r import, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
folder<-'Outputs/'
filenamelist<-list.files(folder, pattern='daily_prediction_results')

SITE_IDlist<-sapply(strsplit(filenamelist, "_"), "[", 3)
PARENT_IDlist<-sapply(strsplit(filenamelist, "_"), "[", 2)

depthfolder<-'Inputs/Sensor_Files/'
depthfilenamelist<-list.files(depthfolder, pattern='csv')
depthPARENT_IDlist<-  sapply(strsplit(depthfilenamelist, "_"), "[", 2)


# Initialize results dataframe
SMresults <- data.frame(matrix(ncol = 13, nrow = 0))
colnames(SMresults) <- c("Parent_ID", "Site_ID", "daysofdata",                      "ERdailymeanmean_gO2/m2day", "GPPdailymeanmean_gO2/m2day",                           "K600dailymeanmean_m/day", "mean_depth_m",
                          "ERdailymeanmean_gO2/m3day","GPPdailymeanmean_gO2/m3day","K600vsERrsq", "K600vsERp", "ER_removed", "GPP_removed")

# Loop through each file
for(i in 1:length(PARENT_IDlist)) { 
  # Load data
  data <- read.csv(paste(folder, filenamelist[i], sep = ''), header = TRUE)
  numdays <- length(data$ER_daily_mean[!is.na(data$ER_daily_mean)])

  # Count initial number of non-NA values for each variable
  initial_counts <- data %>%
    summarise(
      ER_initial_count = sum(!is.na(ER_daily_mean)),
      GPP_initial_count = sum(!is.na(GPP_daily_mean)),
      ER_positive_count = sum(ER_daily_mean > 0, na.rm = TRUE),
      GPP_negative_count = sum(GPP_daily_mean < 0, na.rm = TRUE)

    )

  if (initial_counts$ER_positive_count == numdays|initial_counts$ER_positive_count > 27){
  ER_daily_mean_mean <- mean(data$ER_daily_mean, na.rm = TRUE)
  GPP_daily_mean_mean <- mean(data$GPP_daily_mean, na.rm = TRUE)
  K600_daily_mean_mean <- mean(data$K600_daily_mean, na.rm = TRUE)
  
  removed_counts <- initial_counts %>%
    mutate(
      ER_removed = ER_positive_count,
      GPP_removed = GPP_negative_count) %>%
    dplyr::select(ER_removed, GPP_removed)

  }else{
  # Filter out ER > 0.5 and GPP < -0.5
  filtered_data <- data %>%
    filter(!(ER_daily_mean > 0.5 | GPP_daily_mean < -0.5))

  # Count the number of removed values per variable
  removed_counts <- initial_counts %>%
    mutate(
      ER_removed = ER_initial_count - sum(!is.na(filtered_data$ER_daily_mean)),
      GPP_removed = GPP_initial_count - sum(!is.na(filtered_data$GPP_daily_mean))
    ) %>%
    dplyr::select(ER_removed, GPP_removed)

  # Calculate means for filtered data
  ER_daily_mean_mean <- mean(filtered_data$ER_daily_mean, na.rm = TRUE)
  GPP_daily_mean_mean <- mean(filtered_data$GPP_daily_mean, na.rm = TRUE)
  K600_daily_mean_mean <- mean(filtered_data$K600_daily_mean, na.rm = TRUE)
}
  # Reshape data for ggplot
  data_long <- data %>%
    dplyr::select(ER_daily_mean, GPP_daily_mean, K600_daily_mean) %>%
    pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

  # Plot histograms
  plot <- ggplot(data_long, aes(x = Value, fill = Variable)) +
    geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
    facet_wrap(~ Variable, scales = "free_x") +
    labs(title = paste("Histograms of ER, GPP, and K600 Daily Means for", PARENT_IDlist[i], SITE_IDlist[i]),
         x = "Daily Mean Value",
         y = "Frequency") +
    theme_bw() +
    scale_fill_manual(values = c("ER_daily_mean" = "skyblue", "GPP_daily_mean" = "salmon", "K600_daily_mean" = "lightgreen")) +
    theme(legend.position = "none", aspect.ratio = 1) +
    geom_vline(data = data_long %>% filter(Variable == "ER_daily_mean"), aes(xintercept = 0), color = "blue", linetype = "solid") +
    geom_vline(data = data_long %>% filter(Variable == "GPP_daily_mean"), aes(xintercept = 0), color = "blue", linetype = "solid") +
    geom_vline(data = data_long %>% filter(Variable == "ER_daily_mean"), aes(xintercept = 0.5), color = "red", linetype = "solid") +
    geom_vline(data = data_long %>% filter(Variable == "GPP_daily_mean"), aes(xintercept = -0.5), color = "red", linetype = "solid")

  # Save histogram
  ggsave(filename = paste0("Histogram_Plots/histogram_", PARENT_IDlist[i], "_", SITE_IDlist[i], ".png"), plot = plot, width = 8, height = 6)

  # Calculate depth information
  depthfilenumber <- which(depthPARENT_IDlist == PARENT_IDlist[i])
 depthdata <- read_csv(
  paste0(depthfolder, depthfilenamelist[depthfilenumber]),
  comment = '#',
  na = c('N/A', '', 'NA', '-9999')
)
  meandepthm <- mean(depthdata$Depth, na.rm = TRUE)
  ERdailymeanmeanm3 <- ER_daily_mean_mean / meandepthm
  GPPdailymeanmeanm3 <- GPP_daily_mean_mean / meandepthm

  # Linear regression and statistics
  lmK600ER <- lm(ER_daily_mean ~ K600_daily_mean, data = data)
  K600vsERp <- summary(lmK600ER)$coefficients[2, 4]
  K600vsERrsq <- summary(lmK600ER)$r.squared

SMresults[nrow(SMresults) + 1, ] <- c(
  PARENT_IDlist[i], SITE_IDlist[i], numdays,
  ER_daily_mean_mean, GPP_daily_mean_mean, K600_daily_mean_mean,
  meandepthm, ERdailymeanmeanm3, GPPdailymeanmeanm3,
  K600vsERrsq, K600vsERp,
  removed_counts$ER_removed, removed_counts$GPP_removed
)
}

# Flag as -9999 sites where SM failed or where the overall SM parameters are nonsensical 
SMresults <- SMresults %>%
  mutate_at(
    vars("ERdailymeanmean_gO2/m2day", "GPPdailymeanmean_gO2/m2day", "K600dailymeanmean_m/day", "ERdailymeanmean_gO2/m3day", "GPPdailymeanmean_gO2/m3day", "K600vsERrsq", "K600vsERp"),
    # funs(ifelse(Parent_ID %in% c("SSS001", "SSS007","SSS010","SSS011", "SSS016","SSS022","SSS023","SSS028","SSS033","SSS034","SSS038","SSS040","SSS042","SSS044","SSS045"), -9999, .))
     funs(ifelse(Parent_ID %in% c("SSS023","SSS028","SSS038"), -9999, .))
  )
```
##write CSV outupt
```{r output, echo=FALSE}
write.csv(SMresults, paste(folder,"v2_SSS_combined_SM_results.csv",sep=""), row.names=FALSE)
```
